Using AmCAT codebooks in R 
==========================

Codebooks in AmCAT are essentially hierarchical lists of codes that have one or more labels attached to them. 
We can use this feature to create a query dictionary by adding codes with a 'human' label, for example in English,
and with the keyword query to use. 

The hierarchical nature of codebooks makes it possible to have more general codes (e.g. economy) 
with below them more specific codes (e.g. inflation). 

This howto will show how to download a codebook, use it for querying articles, 
and use the hierarchical nature to classify results.

Downloading codebooks
=====================

The `amcat.gethierarchy` function can be used to download all codes and their hierarchy from a codebook:

```{r, message=F}
library(amcatr)
conn = amcat.connect("http://amcat.nl")
h = amcat.gethierarchy(conn, codebook_id=374, languages=c("en", "query"))
h
```

The resulting data frame has a column `code` that specifies the code and a column `parent` that specifies the parent of the code.
For example, the parent of `nuclear` is `wmd`, which has `goals` as a parent, which does not itself have a parent. 

Querying codes
==============

The functions `amcat.hits` and `amcat.aggregate` described in the Querying howto can be directly used from a codebook:

```{r, message=FALSE}
counts = amcat.aggregate(conn, queries=h$label.query, labels=h$label.en, sets=10271)
counts
```



Adding categories
=================

These parent-child relations define the hierarchy, but they are not very easy to use.
The `amcat.hierarchy.cats` function creates category variables that specify in which part of the tree a code is found:

```{r, message=FALSE}
h = amcat.hierarchy.cats(h, maxdepth=1)
h
```

So, nuclear has 'goals' as its main category (`cat.1`), and 'wmd' as a secondary category. 
This information can be directly merged with the counts data:

```{r, message=F}
counts = merge(h, counts, by.x="code", by.y="query")
counts
tapply(counts$count, counts$cat.1, sum)
```

Note that if the queries have a possible overlap, it is not correct to simply add counts like this. 
It is probably a better idea to get the results per article, and then count the number of unique articles
that contain any of the codes in a specific category: 

```{r, message=F}
arts = amcat.hits(conn, queries=h$label.query, labels=h$label.en, sets=10271)
head(arts)
table(arts$query)
```

As can be seen in the output, `amcat.hits` returns a data frame with counts per article per query.
The tabulation gives the same results as `amcat.aggregate` shown above, as would be expected. 

Now, let's merge the data, and count unique article ids per cat.1 instead:

```{r, message=FALSE}
arts = merge(h, arts, by.x="code", by.y="query")
head(arts)
tapply(arts$id, arts$cat.1, function (x) length(unique(x)))
```

Compared to the earlier results, we can see that 5 goals articles and 2 means articles were indeed overlapping. 
If you plan to do a lot of different plots, categorizations, etc. on the same basic queries and data,
it is very flexible and often more efficient to retrieve all data from AmCAT first, 
and then do the tabulating, filtering, etc. in R rather than by using `amcat.aggregate`.