Automatic Keyword Analysis with the AmCAT API in R
==================================================

This document is an introduction to using the AmCAT API from R. 
The API works using `HTTP GET` requests. 
While it is entirely possible to use the API directly with a library like `RCurl`,
we have made a package `amcatr` to make it easier to use the API, 
see [github.com/amcat/amcat-r](github.com/amcat/amcat-r).

This document is produced using /R Markdown/. 
This means that the original file contains the actual R commands that are processed
and turned into the document you are reading now. 
In this document, R code and output is typeset in shaded boxes with a `monospace` font. 
The output and messages are prefixed by double hash tags (`##`). 
Example:

```{r}
2+2
```

Getting started: installing, connecting and accessing the API
==============


You can install `amcatr` directly from the github repository using devtools:

```{r test, results='hide', tidy=FALSE, message=FALSE, warning=FALSE, cache=TRUE}
if (!require(devtools)) {install.packages("devtools"); library(devtools)}
install_github(repo="amcat-r", username="amcat")
```

Note: Every time you run the `install_github` command, 
the latest version of the package is automatically installed from `github`. 

Connecting to AmCAT
-------------------

Before you can use the AmCAT API, you need to connect to it using `amcat.connect`.
This requests an authentication token from the specified AmCAT server and stores it for further commands.

```{r eval=FALSE}
library(amcatr)
conn = amcat.connect("http://amcat.nl", username='example', passwd='secret')
```

If you don't specify a username or password,
it will use the name of the logged in user and search for a password in a file `.amcatauth` in your home directory.
This file should be a csv file with one line per host, giving host, username, and password: 
`http://amcat.nl,example,secret`. 
When this is in place, you can connect without having to type the password every time 
(and without saving your password in a code file, which is usually a bad idea!): 

```{r, message=FALSE}
library(amcatr)
conn = amcat.connect("http://amcat.nl")
```

Retrieving information from AmCAT
---------------------------------

All functions in the `amcatr` library start with the prefix `amcat.`.  
The basic command to get information from the API is the `amcat.getobjects`.
For example, the following downloads a list of all accessible projects:

```{r}
projects = amcat.getobjects(conn, "projects/")
head(projects[, c("id", "name")])
```

As you can see, the `amcat.getobjects` call is translated into an `HTTP GET` request to the `api/v4/projects` address.
The resource is often specified hierarchically, so the list of sets in project 442 
(a publicly accessible project containing articles from wikinews)
can be queried as follows:

```{r, message=FALSE}
amcat.getobjects(conn, c("projects", 442, "articlesets"))
```

Note that the slash at the end is required for most resources, 
but it is automatically appended if you specify a hierarchical resource such as above. 

Querying AmCAT
==============

The functionality of the query page in the AmCAT navigator is replicated in two `amcatr` functions.
`amcat.hits` gives a list of articles with how often a keyword is found 
(the Article List function on the website)
while `amcat.aggregate` gives total number of articles per query and optionally per date or medium
(the Graph/Table function).

These examples use [Set 10271](http://amcat.nl/navigator/projects/442/articlesets/10271/) 
(in [project 442](http://amcat.vu.nl/navigator/projects/442/articlesets/))
which contains the wikinews articles in the category Iraq. 

Aggregate Queries
-----------------

Using `amcat.aggregate`, 
it is also possible to directly get the amount of articles per query and per time interval.
For example, the code below gets the amount of articles for Obama, Bush and the total articles per year:

```{r}
a = amcat.aggregate(conn, sets=10271, axis1="year",
                    queries=c("*", "obama", "bush"), 
                    labels=c("Total", "Obama", "Bush"))
head(a)
```

As above, this can be transformed from 'long' to 'wide' format using the `cast` function (from the `reshape` package) and plotted:

```{r}
wide = cast(a, year~query, value="count")
plot(wide$year, wide$Bush / wide$Total, type='l', frame.plot=F, ylim = c(0, 0.4), col="red",
     xlab="Year", ylab="% of Articles", main="Percentage of articles mentioning US Presidents")
lines(wide$year, wide$Obama / wide$Total, col="blue")
legend("top",legend=c("Bush", "Obama"), col=c("red", "blue"), lty=1)
```


Getting hits per article
------------------------

The examples above show that it is quite easy to get aggregate data from AmCAT directly. 
Often, however, it is desirable to query the hits per article and use those results, 
for example for computing the co-occurrence of terms or for more flexible filtering and aggregation.

The `amcat.hits` command allows you to get a list of articles matching a query. 
For example, to search for all articles containing the term 'tyrant' in this set, we can use the following

```{r, message=FALSE}
amcat.hits(conn, queries="tyrant", sets=10271)
```

So, this query could be found in two documents, occurring exactly once in each document. 
We can also search for multiple queries simultaneously by specifying a vector of queries,
and we can use the `labels=` argument to specify the result name:

```{r, message=FALSE}
h = amcat.hits(conn, queries=c("tyrant OR brute", "saddam"), labels=c("tyrant", "saddam"), sets=10271)
head(h)
```

As you can see, the `query` column contains the label as specified in the `labels=` argument. 

Adding metadata
---------------

To do something useful with these data, we normally need to add the metadata (date, source, etc.) first. 
The following code retrieves the metadata for set 10271 and adds it to the queries:

```{r, message=FALSE}
meta = amcat.getarticlemeta(conn, set=10271, dateparts=TRUE)
head(meta)
h = merge(h, meta, all.x=TRUE)
head(h)
```

Now, we can plot the results over time, e.g. per year:

```{r, message=FALSE}
library(reshape)
peryear = cast(h, year ~ query, value="count", fun.aggregate=sum)
plot(peryear$year, peryear$saddam, type='l', frame.plot=F, 
     xlab="Year", ylab="Articles", main="Articles mentioning Saddam")
```

The large number of articles in 2013 is most likely caused by wikinews containing more articles that year
rather than by Saddam being more salient in that year rather than earlier. 
Using the metadata, we can add the total number of articles, 
and plot the percentage rather than the absolute number of articles:

```{r, message=FALSE}
total = aggregate(meta$id, by=list(meta$year), FUN=length)
colnames(total) = c("year", "total")
peryear = merge(peryear, total)
plot(peryear$year, peryear$saddam / peryear$total, type='l', frame.plot=F, 
     xlab="Year", ylab="% of Articles", main="Percentage of articles mentioning Saddam")
```

So, quite interestingly it seems that on wikinews, 
Saddam was mentioned more often in 2013, even relative to the total amount of articles.
(This probably says more about wikinews than about the state of the world)

